Stage-Splitting a Modal Language

### 2014-09-28 paper/semantics.tex
Wrote a simple section 3.0 and 3.1

As stated in \cref{sec:staging}, we operate on a language called \lang,
which is a typed two-stage lambda calculus featuring products, sums, and isorecursive types.

\subsection{Statics}
The grammar and type system of \lang\ are given in \ref{fig:grammar,fig:statics}. 
They are a simple adaptation of those of \cite{davies96}, 
restricted to two stages\footnote{This restriction is not a big deal...} and extended with general sums and recursion.

Typing judgments and context variables are annotated with stages after an $@$ symbol.
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.

\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo.

The stage-\bbone\ type $\fut A$ contains encapsulated
stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
opaquely by stage-\bbone\ code, as (by the requirements of staging) they cannot be evaluated until stage \bbtwo.
While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
That is, there is a one-way dependence between stages at the term level.

Alternatively, the stages are mutually dependent at the term level.
The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  $\next$ and $\prev$ are the
only ways in \lang\ to alter the stage of a term; we surround their arguments
with braces in \lang\ syntax to clearly indicate stage boundaries within a
program.

%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality,\cite{cave14}.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
It is possible to implement $\pause$ from other \lang\ features, but
we provide it as a core primitive to simplify our examples.

Recall the fast exponent function from above.
We present a valid staging of it here:
\begin{lstlisting} 
let fexp (b : $int, p : int) : $int =
	if p == 0 then
		next{1}
	else if (p mod 2) == 0 then
		next{let x = prev{fexp(b,p/2)} in x*x}
	else
		next{prev{b} * prev{fexp(b,p-1)}}		
\end{lstlisting}

\TODO explain why we staged this the way we did

The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns a stage-\bbtwo\ result. 
The {\tt if} predicates and exponent decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.

>>>>

As stated in \cref{sec:staging}, we operate on a language called \lang,
which is a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
The grammar of \lang\ can be found in \cref{fig:grammar}.

\subsection{Statics}
The grammar and type system of \lang\ are a simple adaptation of those of \cite{davies96}, 
restricted to two stages\footnote{This restriction to two stages is made primarily for 
simplicity of presentation.  All of these techniques could feasibly be extended to more
complicated stage systems.} and extended with general sums and recursion.

As explained before, the $\fut$ constructor is a way to encapsulate a stage \bbtwo\ type at stage \bbone.  
But because of the two-stage assumption, $\fut$ has no meaning when used at stage \bbtwo.
This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
This context is only augmented by recursive types.

The typing judgment, like type validity, is annotated with a stage.
Thus we can read $\typeswor x A$ as saying that term $x$ has type $A$ at world $w$ under the context $\Gamma$.
The typing rules are given in \cref{fig:statics}.
The staging features ($\next$, $\prev$, and $\pause$) behave as described in the previous section,
and all other features behavior in the standard way and are essentially agnostic to stage. 
Variables in the context are annotated with the stage at which they were introduced 
and can only be used at the same stage.
As one would expect, the typing judgment only produces well-formed types 
(by induction on the derivation of the typing judgment).
